---
title: "Assignment 3: Data Exploration"
author: "Julia Weinberg, Section #01"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**
```{r message=FALSE}
getwd()
#install.packages(tidyverse)
library(tidyverse)
#install.packages(lubridate)
library(lubridate)

Neonics <- read.csv("../Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE) 
#upload Neonics

Litter <- read.csv("../Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE) 
#upload Litter


```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Neonicotinoids have been shown to have adverse impacts on pollinators. A reduction in pollinator species would lead to food scarcity on the planet, so assessing the ecotoxicology of neonicotinoids would be beneficial to protect insect species, not only limited to pollinators. 

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: We might be interested in studying leaf litter and woody debris because it allows us to see the organic content that will decompose into the top level of the soil. If the leaf litter is high in nutrients, then the soil will be nutrient dense and be able to support a thriving ecosystem. 

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
*Ground traps are sampled yearly.
*Elevated traps are sampled biweekly in deciduous forests and every 1-2 months in evergreen sites.
*Elevated traps in deciduous locations are often not sampled for up to six months during the dormant winter seasons. 



## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
dim(Neonics) #get dimensions of Neonics


```

6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
summary(Neonics$Effect) #get summary of Neonics Effect column

```

> Answer: 
The most commonly studied effects are mortality, behavior, feeding behavior, and population. These are of specific interest becasue they give insight into the survival rate of insect species as a result of the pesticide. Since species extinction is something that might be of concern, these are especially important parameters to look at.  
7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics$Species.Common.Name) #summary of Neonics species 
```

> Answer: The most commonly studied species include: Honey Bee, Parasitic Wasp, Buff Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee, and Italian Honeybee. All of these species are polinators and they are of increased interest because they are necessary for plant cultivation. 

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.) #determining class of Conc.1..Author

```

> Answer:Conc.1..Author is factor data. It is not numeric because it is representing categorical data. 



## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}
ggplot (Neonics, aes(x = Publication.Year)) +
  geom_freqpoly()
#create plot of Publication.Year
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}

ggplot(Neonics, aes(x = Publication.Year, color = Test.Location)) + geom_freqpoly()
#create plot of Publication.Year color coded by Test.Location
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are lab and field natural. Field natural in most common in 2010 and lab is most common in 2015.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar() #create bar graph of endpoint frequency
summary(Neonics$Endpoint) #improve readability of axes

```

> Answer: 
> The two most common end points are LOEL and NOEL. LOEL is defined as the lowest observable effect level, meaning that it was the lowest dose that produced effects. NOEL is defined as the no observable effect level, meaning that it was the highest dose given where no effects were produced. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
class(Litter$collectDate) #determine class of collectDate

Litter$collectDate <- as.Date(Litter$collectDate)
#change collectDate to date class


unique(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
df_unique <- unique(Litter$plotID)
length(df_unique) #determine how many plots sampled at Niwot Ridge

summary(df_unique) #summary of plots sampled at Niwot Ridge
```

> Answer: Unique shows the sum of plots at the Niwot Ridge area, however summary shows how many individual counts there are of Niwot Ridge.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}
ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar() #create bar graph of litter on Niwot Ridge
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}
ggplot(Litter, aes(x = dryMass, y = functionalGroup)) +
 geom_boxplot() #create boxplot of litter on Niwot Ridge
ggplot(Litter, aes(x = dryMass, y = functionalGroup)) +
  geom_violin() #create violin plot of litter on Niwot Ridge
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The box plot is more effective than the violin plot because it shows the mean as well as the quartile range. The violin plot alternitively just shows the entire range of results which is not as indicative for what is commonly observed. 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: The litter that tends to have the highest biomass is needles. 