---
output:
  pdf_document: default
  html_document: default
---
---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Julia Weinberg"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1 
getwd()
library(tidyverse)
library(agricolae)
library(lubridate)

NTL_LTER <- read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE, header = TRUE)

NTL_LTER$sampledate <- as.Date(NTL_LTER$sampledate , format = "%m/%d/%y") 
#format as date function
#2

my.theme <- theme_gray(base_size = 14) + #set base theme and size
  theme(axis.text = element_text(color = "darkblue"),  #color text dark blue
        legend.position = "bottom") #align legend on bottom
theme_set(my.theme) #set theme
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no relationship between depth and lake temperature 
recorded in July.
Ha: There is a relationship between lake temperature 
and depth recorded in July.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

July.Temps <- NTL_LTER %>%
  mutate(month = month(sampledate))%>% #add column for month
  filter(month == "7") %>% #only include dates with July as month
  select(lakename, year4, daynum, depth, temperature_C)%>% #select variables
filter(!is.na(temperature_C)) #filter out NA's



#5
temp.by.depth <- ggplot(July.Temps, aes(x = depth, y = temperature_C)) + 
  #plot temp by depth
  geom_point()+
  ylim(0, 35)+ #limit data plotted 
  geom_smooth(method = "lm", color = "purple")+ #add best fit line
my.theme #include my theme
print(temp.by.depth) #print graph

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: This plot shows that as depth increases, temperature tends to decrease, therefore pointing to an inverse relationship between the two variables. However, this relationship is not entirely linear as temperature decreases begin to level off as depth continues to increase. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
temp.depth.reg <- lm(data = July.Temps, temperature_C ~ depth) 
#linear regression of july temperature by lake depth
summary(temp.depth.reg) #summary of regrerssion

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The R-squared value is 0.7387, which is closer to one, which would mean that the two variables are perfectly correlated. The p-value is 2.2e-16, which is much smaller than 0.05, therefore meaning that we should reject the null hypothesis, showing that there is some relationship between temperature and depth. The estimate for depth is -1.95, meaning that for every increase in depth of 1m, temperature decreased by 1.95 degrees C. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
library(dplyr)

TEMP.regression <- lm(data = July.Temps, temperature_C ~ depth + 
              year4 + daynum) 
#create regression of temperature with chosen variables

step(TEMP.regression) #get step values


#10


TEMP.regression.model <- lm(data = July.Temps, temperature_C ~ depth + year4 + 
daynum)
summary(TEMP.regression.model) #get summary of temperature by chosen variables

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of variables is depth, daynum, and year. The AIC suggests that using depth, year, and daynum provides a better correlation to temperature than just depth alone. When depth was used alone, only 73.87% of variance in lake temperature was accounted for, however when year and daynum are added, 74.11% of the variance in temperature is accounted for. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
july.temp.anova <- aov(data = July.Temps , temperature_C ~ lakename)
summary(july.temp.anova) #create anova of July temps by chosen variables

july.temp.linear <- lm(data = July.Temps, temperature_C ~ lakename)
summary(july.temp.linear) 
#create linear regression of July temps by chosen variables


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant differnce between all of the lakes as they all have p-values less than 0.05.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

plot.of.temp.by.depth <- ggplot(July.Temps, aes(x = depth, y = temperature_C, color = lakename))+
  geom_point(alpha = 0.5)+ #set transparancy to 50%
  ylim(0, 35)+ #limit data plotted 
  geom_smooth(method = "lm", se = FALSE)+#add best fit line
my.theme #include by theme
print(plot.of.temp.by.depth) #print graph
 
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(july.temp.anova) #use Tukey's HSD test

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul lake is the only one with a similar temperature. Its difference in temperature mean is closest to Peter Lake at only -0.49, however it is not a statistically significant similarity as its p-value is greater than 0.05. Central Long Lake is most distinct form the other lakes as it has the greatest difference in mean temperature comapred to the other lakes in the data set. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: A HSD test would be good to use as it would group lakes that are similar to one another by letter categories. 


